---
title: "Branching Processes and Monte Carlo Simulation"
author: "Brett Hallum, Chris Ficklin, and Ryan Shuhart"
date: "April 2017"
output: html_document
runtime: shiny
subtitle: MSDS 7333-401
---
https://github.com/rlshuhart/Unit-12_Simulation_Study_of_a_Branching_Process/
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Q.6 Figure 7.4 is a custom visualization of the birth and completion times for a tree. Design an alternative custom visualization of the return value from familyTree().

## Introduction 
Monte Carlo simulations can be used to test a theory by taking many random samples approximating characteristics. 

## Background 


## Methods
The familyTree simulation determines how many children are born from each parent and will continue to simulate until either no children are born or the max generations parameter is reached

## Results
The following are vizualizations of the birth and completion times for a tree using the same data from Fig 7.4 in <i><b>Data Science in R</i></b> by Deborah Nolan and Duncan Temple Lang. Figure 1 is an alternative view as a downward facing tree. In Figure 2, the design is mostly the same as used in the book, however, it is interactive and lambda, kappa, max generation, and random seed are adjustable.

### Alternative Vizualization of a Family Tree Shown in Fig 7.4 of <i><b>Data Science in R</i></b>
```{r, echo=FALSE, fig.width=10, message=FALSE, warning=FALSE}
# Only Simulation Code
genKidsV = function(bTimes, cTimes, parentID, lambda = 0.5, kappa = 0.3) {
  # Determine how many children each job has
  parentAge = cTimes - bTimes
  numKids = rpois(n = length(parentAge), lambda = lambda * parentAge)
  
  if (sum(numKids) == 0) return(NULL)
  
  # Determine the birth times of the children  
  kidStats = 
    mapply(function(n, min, max) {
      births = sort(runif(n, min, max))
      runtimes = rexp(n, rate = kappa)
      completes = rep(max, n) + runtimes
      data.frame(births, completes)
    },
    n = numKids , min = bTimes, max = cTimes, 
    SIMPLIFY = FALSE)
  
  
  return(data.frame(parentID = rep(parentID, numKids),
                    kidID = 1:sum(numKids), 
                    births = unlist(lapply(kidStats, "[[", "births")), 
                    completes = unlist(lapply(kidStats,"[[", "completes"))
  ))
}

familyTree = function(lambda = 0.5, kappa = 0.3, 
                      maxGen = 10, maxOffspring = 1000) {
  
  # Return value - a list with 1 data frame per generation.
  allGens = vector(mode = "list", length = maxGen)
  
  # Generate root of the tree
  allGens[[1]] = data.frame(parentID = NA, kidID = 1, 
                            births = 0, 
                            completes = rexp(1, rate = kappa))
  
  currentNumOffspring = 0
  
  # Generate future generations, one at a time.
  for (i in 2:maxGen) {
    nextGen = genKidsV(bTimes = allGens[[ (i - 1) ]]$births,
                       cTimes = allGens[[ (i - 1) ]]$completes,
                       parentID = allGens[[ (i - 1) ]]$kidID,
                       lambda = lambda, kappa = kappa)
    if (is.null(nextGen)) return(allGens[ 1:(i - 1) ]) 
    allGens[[ i ]] = nextGen
    currentNumOffspring = currentNumOffspring + nrow(nextGen)
    if (currentNumOffspring > maxOffspring) 
      return(allGens[1:i])
  }  
  allGens
}

draw_tree <- function(lambda = .4, kappa = 1, maxGen = 10, seedx=12062013){
  library(dplyr)
  library(ggplot2)
  #lambda = .4; kappa = 1; maxGen = 10; seedx=12062013
  set.seed(seedx); tree <- familyTree(lambda=lambda, kappa = kappa, maxGen = maxGen)
  
  tree_df <- data.frame()
  for (i in 1:length(tree)){
    # convert gen i to data frame
    tmp <- as.data.frame(tree[i])
    # lable gen i
    tmp$Gen <- i
    tree_df <- rbind(tree_df, tmp)
  }
  
  # Add y cord
  #tree_df$x_cord <- 1:nrow(tree_df)*.5
  tree_df$x_cord <- c(0.0, -1.0, .5, -.5, -1.5, 1.5, -2.0, 1.0, 2.0)
  
  # Add parent generation reference
  tree_df$Parent_Gen <- tree_df$Gen-1
  
  # Self join the table to identify the births on the parent
  marks <- inner_join(select(tree_df, kidID, Gen, x_cord), 
                      select(tree_df, parentID, Parent_Gen, births), 
                      by=c("kidID" = "parentID", "Gen"="Parent_Gen"))
  
  # Connect birth to life line
  conn_lines <- inner_join(select(tree_df, births, x_cord, Gen),
                           select(marks, births, x_cord), by="births")
  
  # Generation separation locations
  gen_lines <- tree_df %>% 
    group_by(Gen) %>% 
    summarize(Gen_Break = max(x_cord)+.25) %>%
    mutate(Gen_Label = paste("Gen", Gen))
  
  g <- ggplot(tree_df) + 
    geom_segment(aes(y = -births, x = x_cord, yend = -completes, xend = x_cord, color=as.factor(Gen-1)), size=1.25) +
    geom_segment(aes(y = -births, x = x_cord.x, yend = -births, xend = x_cord.y, color=as.factor(Gen-1)), data = conn_lines,  size=1.25) +
    geom_point(aes(y = -births, x = x_cord), data = marks, shape=19, size=3, color="#007F00") +
    scale_color_discrete(name="Generation")+ 
    theme_bw() + 
    labs(title = "Fig. 1") +
    theme(axis.title.y=element_blank(),
          axis.title.x=element_blank(),
          panel.grid.major.y = element_blank(),
          panel.grid.minor.y = element_blank(),
          panel.grid.major.x = element_blank(),
          panel.grid.minor.x = element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank(),
          axis.text.y=element_blank(),
          axis.ticks.y=element_blank()) 
  
  return(g)
}

draw_tree()

```


### Interactive Visualization of Fig 7.4 in <i><b>Data Science in R</i></b>
```{r tabsets, echo=FALSE}

ui <- fluidPage(
  #headerPanel('Simulated Branch Process Family Tree'),
  sidebarPanel(
    sliderInput("lambda", "Select Lambda", min = .1, max=1.2, value=.4),
    sliderInput("kappa", "Select Kappa", min = .1, max=1.5, value=1),
    sliderInput("MaxGen", "Select Max Generations", min = 1, max=20, value=10),
    numericInput("seed", "Input Seed", value = 12062013)
    ),
  mainPanel(
    plotOutput("GenTree", height = "800px")
  )
)

server <- function(input, output) {

  vis <- renderPlot({

    genKidsV = function(bTimes, cTimes, parentID, lambda = 0.5, kappa = 0.3) {
      # Determine how many children each job has
      parentAge = cTimes - bTimes
      numKids = rpois(n = length(parentAge), lambda = lambda * parentAge)
      
      if (sum(numKids) == 0) return(NULL)
      
      # Determine the birth times of the children  
      kidStats = 
        mapply(function(n, min, max) {
          births = sort(runif(n, min, max))
          runtimes = rexp(n, rate = kappa)
          completes = rep(max, n) + runtimes
          data.frame(births, completes)
        },
        n = numKids , min = bTimes, max = cTimes, 
        SIMPLIFY = FALSE)
      
      
      return(data.frame(parentID = rep(parentID, numKids),
                        kidID = 1:sum(numKids), 
                        births = unlist(lapply(kidStats, "[[", "births")), 
                        completes = unlist(lapply(kidStats,"[[", "completes"))
      ))
    }
    
    familyTree = function(lambda = 0.5, kappa = 0.3, 
                          maxGen = 10, maxOffspring = 1000) {
      
      # Return value - a list with 1 data frame per generation.
      allGens = vector(mode = "list", length = maxGen)
      
      # Generate root of the tree
      allGens[[1]] = data.frame(parentID = NA, kidID = 1, 
                                births = 0, 
                                completes = rexp(1, rate = kappa))
      
      currentNumOffspring = 0
      
      # Generate future generations, one at a time.
      for (i in 2:maxGen) {
        nextGen = genKidsV(bTimes = allGens[[ (i - 1) ]]$births,
                           cTimes = allGens[[ (i - 1) ]]$completes,
                           parentID = allGens[[ (i - 1) ]]$kidID,
                           lambda = lambda, kappa = kappa)
        if (is.null(nextGen)) return(allGens[ 1:(i - 1) ]) 
        allGens[[ i ]] = nextGen
        currentNumOffspring = currentNumOffspring + nrow(nextGen)
        if (currentNumOffspring > maxOffspring) 
          return(allGens[1:i])
      }  
      allGens
    }
    
    draw_tree <- function(lambda = .4, kappa = 1, maxGen = 10, seedx=12062013){
      library(dplyr)
      library(ggplot2)
      set.seed(seedx); tree <- familyTree(lambda=lambda, kappa = kappa, maxGen = maxGen)
      
      tree_df <- data.frame()
      for (i in 1:length(tree)){
        # convert gen i to data frame
        tmp <- as.data.frame(tree[i])
        # lable gen i
        tmp$Gen <- i
        tree_df <- rbind(tree_df, tmp)
      }
      
      # Add y cord
      tree_df$y_cord <- 1:nrow(tree_df)*.5
      
      # Add parent generation reference
      tree_df$Parent_Gen <- tree_df$Gen-1
      
      # Self join the table to identify the births on the parent
      marks <- inner_join(select(tree_df, kidID, Gen, y_cord), 
                          select(tree_df, parentID, Parent_Gen, births), 
                          by=c("kidID" = "parentID", "Gen"="Parent_Gen"))
      
      # Connect birth to life line
      conn_lines <- inner_join(select(tree_df, Gen, births, y_cord),
                               select(marks, births, y_cord), by="births")
      
      # Generation separation locations
      gen_lines <- tree_df %>% 
        group_by(Gen) %>% 
        summarize(Gen_Break = max(y_cord)+.25) %>%
        mutate(Gen_Label = paste("Gen", Gen))
      
      g <- ggplot(tree_df) +
        geom_segment(aes(x = births, y = y_cord, xend = completes, yend = y_cord, color=as.factor(Gen-1)), size=1.25) +
        geom_segment(aes(x = births, y = y_cord.x, xend = births, yend = y_cord.y, color=as.factor(Gen-1)), data = conn_lines, linetype = 3) +
        geom_point(aes(x = births, y = y_cord), data = marks, shape=19, size=3, color="#007F00") +
        geom_hline(aes(yintercept=Gen_Break), data = gen_lines, linetype = 2) +
        scale_y_continuous(labels=gen_lines$Gen_Label, breaks=gen_lines$Gen_Break) +
        theme_bw() +
        labs(x = "Time", title="Fig. 2") +
        theme(axis.title.y=element_blank(),
              panel.grid.major.y = element_blank(),
              panel.grid.minor.y = element_blank(),
              panel.grid.major.x = element_blank(),
              panel.grid.minor.x = element_blank()) +
    scale_color_discrete(name="Generation")
      
      return(g)
    }
    
    lambda <- input$lambda
    kappa <- input$kappa
    maxGen <- input$MaxGen
    seedx <- input$seed
    
    draw_tree(lambda = lambda, kappa = kappa, maxGen = maxGen, seedx=seedx)
    })
  
  output$GenTree <- vis
  
}

shinyApp(ui = ui, server = server, options = list(width = "100%", height = 900))
```

## Conclusion


## Bibliography
* Data Science in R: A Case Studies Approach to Computational Reasoning and Problem Solving by Deborah Nolan and Duncan Temple Lang 
* http://docs.ggplot2.org/current/ 
* https://shiny.rstudio.com/tutorial/