--
title: "Branching Processes and Monte Carlo Simulation"
author: "Brett Hallum, Chris Ficklin, and Ryan Shuhart"
date: "April 2017"
output:
  html_notebook: default
  html_document: default
  pdf_document: default
subtitle: MSDS 7333-401
---
https://github.com/rlshuhart/Unit-12_Simulation_Study_of_a_Branching_Process/

# Q.6 Figure 7.4 is a custom visualization of the birth and completion times for a tree. Design an alternative custom visualization of the return value from familyTree().


```{r}
#Generates kids when passed vectors of birth times and completion times
genKidsV = function(bTimes, cTimes, parentID, lambda = 0.5, kappa = 0.3){
  
  parentAge = cTimes - bTimes
  numKids = rpois(n = length(parentAge), lambda = lambda*parentAge)
  
  if (sum(numKids) == 0){
    return(NULL)
  }
  
  kidStats = mapply(function(n, min, max){
    births = sort(runif(n, min, max))
    runtimes = rexp(n, rate = kappa)
    completes = rep(max, n) + runtimes
    data.frame(births, completes)
  }, n = numKids, min = bTimes, max = cTimes, SIMPLIFY = FALSE)
  
  return (data.frame(parentID = rep(parentID, numKids),
                    kidID = 1:sum(numKids), 
                    births = unlist(lapply(kidStats, "[[", "births")), 
                    completes = unlist(lapply(kidStats,"[[", "completes"))
                    ))
}

#Creates a family tree with a max number of generations and offspring
familyTree = function(lambda=0.5, kappa=0.3, maxGen=10, maxOffspring=1000){
  
  allGens = vector(mode="list", length=maxGen)
  allGens[[1]] = data.frame(parentID = NA, kidID = 1, births = 0, completes = rexp(1, rate = kappa))
  
  currentNumOffspring = 0
  
  for (i in 2:maxGen){
    nextGen = genKidsV(bTimes = allGens[[ ( i - 1 )]]$births,
                       cTimes = allGens[[ ( i - 1 )]]$completes,
                       parentID = allGens[[ ( i - 1 )]]$kidID,
                       lambda = lambda, kappa = kappa)
    
    if (is.null(nextGen)){
      return (allGens[ 1:(i - 1)])
    }
    
    allGens[[i]] = nextGen
    currentNumOffspring = currentNumOffspring + nrow(nextGen)
    
    if (currentNumOffspring > maxOffspring){
      return (allGens[1:i])
    }
  }
  
  allGens
}
```

```{r}
#testing family tree call
seed = 1425

set.seed(seed)

tree = familyTree()
tree

sapply(tree, nrow)
```

```{r}
#prints the number of generations in a family tree and number of total jobs run (children)
exptOne = function(l, k, mG, mO){
  aTree = familyTree(lambda = l, kappa = k, maxGen = mG, maxOffspring = mO)
  
  numGen = length(aTree)
  numJobs = sum(sapply(aTree, nrow))
  
  c(numGen, numJobs)
}

set.seed(seed)
#lambda 1 and kappa 0.5
exptOne(1, 0.5, 10, 1000)
```

```{r, cache=TRUE}
#iterates over exptOne() for a given list of params (various lambda and kappas)
# --TAKES LONG TIME TO RUN DEPENDING ON PARAMS SIZE & REPEATS
monteCarlo = function(params, repeats = 5, mG = 10, mO = 1000){
  
  n = nrow(params)
  
  mcResults = vector("list", length = n)
  
  for (i in 1:n){
    cat("param set is ", i, "\n")
    mcResults[[i]] = replicate(repeats, exptOne(l = params[i,1], k = params[i,2], mG = mG, mO = mO))
  }
  
  mcResults
}

lambdas = c(seq(0.1, 0.6, by=0.1), seq(0.8, 2, by=0.2), seq(2.25, 3, by=0.25))
kappas = c(lambdas, 3.25, 3.5, 3.73, 4, 4.5, 5)

paramGrid = as.matrix(expand.grid(lambdas, kappas))

numRepeats = 200

mcGrid = monteCarlo(params = paramGrid, repeats = 200, mG = 10, mO = 1000)
```
```{r}
rowMeans(mcGrid[[9]])
```
```{r}
#Average generation size and children count for each combination of kappa and lambda above for 200 trials
matSize = length(mcGrid)
rowMeanSum = 0

for (i in 1:matSize){
  cat ("lambda =", paramGrid[i,]["Var1"])
  cat (" kappa =", paramGrid[i,]["Var2"])
  print (rowMeans(mcGrid[[i]]))
  rowMeanSum = rowMeanSum + rowMeans(mcGrid[[i]])
}
```
```{r}
#Overall Average of all runs
rowMeanSum/matSize
```
