---
title: "R Notebook"
output: html_notebook
---
6, 9, 10, or 11

Q.6 Figure 7.4 is a custom visualization of the birth and completion times for a tree. Design
an alternative custom visualization of the return value from familyTree().

Q.9 The branching process was summarized by two statistics: the number of generations
and the number of offspring (see Section 7.9). Consider other summary statistics for the
process. Incorporate them into exptOne(). Carry out a simulation study and create a
visualization of the simulation that uses these additional statistics. Do they confirm the
earlier findings? Do they offer any new insights?

Q.10 Carry out a simulation study to see if the re-parameterization suggested in Section
7.9.1 is appropriate. For example, fix kappa to be 1, and run the simulation for various
values of lambda. Compare the results to other simulations where kappa is c not equal 1, but the ratio of lambda/c matches one of the lambda values from the earlier simulation when kappa was 1.

Q.11 Consider other probability functions to describe the lifetime of a process. Revise familyTree() (see Section 7.8) and genKidsV() (see Section 7.5) to take as an argument the
random number generator for any probability distribution. The functions familyTree()
and genKidsV() are to use this probability distribution (with arguments that may be
specific to the distribution) to generate the completion times of the jobs.

```{r}
#Generates kids when passed vectors of birth times and completion times
genKidsV = function(bTimes, cTimes, parentID, lambda = 0.5, kappa = 0.3){
  
  parentAge = cTimes - bTimes
  numKids = rpois(n = length(parentAge), lambda = lambda*parentAge)
  
  if (sum(numKids) == 0){
    return(NULL)
  }
  
  kidStats = mapply(function(n, min, max){
    births = sort(runif(n, min, max))
    runtimes = rexp(n, rate = kappa)
    completes = rep(max, n) + runtimes
    data.frame(births, completes)
  }, n = numKids, min = bTimes, max = cTimes, SIMPLIFY = FALSE)
  
  return (data.frame(parentID = rep(parentID, numKids),
                    kidID = 1:sum(numKids), 
                    births = unlist(lapply(kidStats, "[[", "births")), 
                    completes = unlist(lapply(kidStats,"[[", "completes"))
                    ))
}

#Creates a family tree with a max number of generations and offspring
familyTree = function(lambda=0.5, kappa=0.3, maxGen=10, maxOffspring=1000){
  
  allGens = vector(mode="list", length=maxGen)
  allGens[[1]] = data.frame(parentID = NA, kidID = 1, births = 0, completes = rexp(1, rate = kappa))
  
  currentNumOffspring = 0
  
  for (i in 2:maxGen){
    nextGen = genKidsV(bTimes = allGens[[ ( i - 1 )]]$births,
                       cTimes = allGens[[ ( i - 1 )]]$completes,
                       parentID = allGens[[ ( i - 1 )]]$kidID,
                       lambda = lambda, kappa = kappa)
    
    if (is.null(nextGen)){
      return (allGens[ 1:(i - 1)])
    }
    
    allGens[[i]] = nextGen
    currentNumOffspring = currentNumOffspring + nrow(nextGen)
    
    if (currentNumOffspring > maxOffspring){
      return (allGens[1:i])
    }
  }
  
  allGens
}
```

```{r}
#testing family tree call
seed = 1425

set.seed(seed)

tree = familyTree()
tree

sapply(tree, nrow)
```

```{r}
#prints the number of generations in a family tree and number of total jobs run (children)
exptOne = function(l, k, mG, mO){
  aTree = familyTree(lambda = l, kappa = k, maxGen = mG, maxOffspring = mO)
  
  numGen = length(aTree)
  numJobs = sum(sapply(aTree, nrow))
  
  c(numGen, numJobs)
}

set.seed(seed)
#lambda 1 and kappa 0.5
exptOne(1, 0.5, 10, 1000)
```

```{r, cache=TRUE}
#iterates over exptOne() for a given list of params (various lambda and kappas)
# --TAKES LONG TIME TO RUN DEPENDING ON PARAMS SIZE & REPEATS
monteCarlo = function(params, repeats = 5, mG = 10, mO = 1000){
  
  n = nrow(params)
  
  mcResults = vector("list", length = n)
  
  for (i in 1:n){
    cat("param set is ", i, "\n")
    mcResults[[i]] = replicate(repeats, exptOne(l = params[i,1], k = params[i,2], mG = mG, mO = mO))
  }
  
  mcResults
}

lambdas = c(seq(0.1, 0.6, by=0.1), seq(0.8, 2, by=0.2), seq(2.25, 3, by=0.25))
kappas = c(lambdas, 3.25, 3.5, 3.73, 4, 4.5, 5)

paramGrid = as.matrix(expand.grid(lambdas, kappas))

numRepeats = 200

mcGrid = monteCarlo(params = paramGrid, repeats = 200, mG = 10, mO = 1000)
```
```{r}
rowMeans(mcGrid[[9]])
```
```{r}
#Average generation size and children count for each combination of kappa and lambda above for 200 trials
matSize = length(mcGrid)
rowMeanSum = 0

for (i in 1:matSize){
  cat ("lambda =", paramGrid[i,]["Var1"])
  cat (" kappa =", paramGrid[i,]["Var2"])
  print (rowMeans(mcGrid[[i]]))
  rowMeanSum = rowMeanSum + rowMeans(mcGrid[[i]])
}
```
```{r}
#Overall Average of all runs
rowMeanSum/matSize
```
